\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-07 k-planar.tex
Pfffft

\begin{abstract}
Dujmovi\'c~et~al.~[\emph{J.~ACM}~'20] recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. Analogous results were obtained for graphs of bounded Euler genus or apex-minor-free graphs. These tools have been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and adjacency labelling. This paper proves analogous product structure theorems for various non-minor-closed classes. One noteable example is $k$-planar graphs (those with a drawing in the plane in which each edge is involved in at most $k$ crossings). We prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. This leads to analogous results for certain types of map graphs, string graphs, graph powers, and nearest neighbour graphs. 
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

@@

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20}.

\begin{thm}[\citep{DJMMUW20}]
	\label{PlanarProduct}
	Every planar graph is a subgraph of:
	\begin{compactenum}[(a)]
		\item $H\boxtimes P$ for some graph $H$ of treewidth at most $8$ and for some path $P$,
		\item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

\note{DW}{In 	\cref{PlanarProduct} and elsewhere, we really should write, ``Every planar graph is isomorphic to a subgraph of''. One alternative is to define that $H$ is \defin{contained} in $G$ if $H$ is isomorphic to a subgraph of $G$. I suggest the latter.}

\note{DW}{I suggest we replace $8$ by $6$ here, and replace ``The starting point for this work is the following `product structure theorem' for planar graphs by \citet{DJMMUW20} (improved by \citet{UWY})''. Okay?}

Here $\boxtimes$ is the strong product\footnote{The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if
	$v=w$ and $xy\in E(B)$, or
	$x=y$ and $vw\in E(A)$, or
	$vw\in E(A)$ and $xy\in E(B)$.}, 
and treewidth\footnote{For a tree $T$, a \defin{$T$-decomposition} of a graph $G$ is a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $T$ such that
(i) for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and
(ii) for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected. The \defin{width} of $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  A \defin{tree-decomposition} is a $T$-decomposition for any tree $T$. The \defin{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.} is an invariant that measures how `tree-like' a given graph is; see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarProduct} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded treewidth to be generalised for planar graphs (with different constants).

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).
\item \citet{DFMS21} use it to make dramatic improvements to the best known bounds for $p$-centered colourings of planar graphs. 
\item \citet{bonamy.gavoille.ea:shorter} use it to find shorter adjacency labellings of planar graphs (improving on a sequence of results going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}), and \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs.
\item The result of \citet{DEJGMM21} implies that, for every integer
$n>0$, there is a `universal' graph $U_n$ with $n^{1+o(1)}$ vertices such that every $n$-vertex planar graph is an induced
subgraph of $U_n$.
\end{compactitem}

\note{DW}{Delete `from 1992' and  `from 2007'? replace ``going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}'' by 
	``starting with the work of \citet{kannan.naor.ea:implicit}''}

\note{DW}{Add \citep{EJM} to the story.}

@@

@@

\note{DW}{Use $\SS$ for shortcut system, okay?}

@@

\begin{thm}[\citep{DJMMUW20}]
	\label{GenusProduct}
	Every graph of Euler genus $g$ is a subgraph of:
	\begin{compactenum}[(a)]
		\item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+8$  and some path $P$.
		\item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$.
	\end{compactenum}
\end{thm}

\note{DW}{I suggest we replace $2g+8$ by $2g+6$ here and replace 4 by 3, and cite \citep{DHHW,UWY}. This will improve the bounds and (more importantly) simplify the presentation of various results (such as \cref{StringPartition,PowerGenus}) which currently distinguish the $g=0$ case. }

@@

  Next, consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.
  % To see that $(B_x:x\in V(T))$ satisifies \tref{ancestor-edge}, observe that,
  % if $xy\in E(H)$, then at least one of $x$ or $y$ is contained in $B_z$ for every node $z$ on the path from $x$ to $y$ in $T$.  If neither $x$ nor $y$ is an ancestor of the other, then some node $z$ on this path has $T$-depth less than that of $x$ and $y$.  If $x\in B_z$ this contradicts the fact that $x$ is the root of $T[x]$.  If $y\in B_z$ this contradicts the fact that $y$ is the root of $T[y]$.
\end{proof}

\referee{1}{6. page 6, last paragraph of Lemma 2: This isn’t quite right since you
have to be careful with $x_0$.}

@@

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  Let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

\referee{2}{Page 8.  Can you add a brief explanation why $\mathcal{S}$ is a
	partition of $V(G)$?  It is clear that the sets in $\mathcal{S}$ are
	disjoint, but why do they cover $V(G)$?}

>>>>

\begin{abstract}
Dujmovi\'c~et~al.~[\emph{J.~ACM}~'20] recently proved that every planar graph is isomorphic to a subgraph of the strong product of a graph of bounded treewidth and a path. Analogous results were obtained for graphs of bounded Euler genus or apex-minor-free graphs. These tools have been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and adjacency labelling. This paper proves analogous product structure theorems for various non-minor-closed classes. One noteable example is $k$-planar graphs (those with a drawing in the plane in which each edge is involved in at most $k$ crossings). We prove that every $k$-planar graph is isomorphic to a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. This leads to analogous results for certain types of map graphs, string graphs, graph powers, and nearest neighbour graphs.
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

@@

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.  We say that a graph $G$ is \emph{contained in} a graph $X$ if $G$ is isomorphic to a subgraph of $X$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (improved by \citet{UWY}).

\begin{thm}[\citep{DJMMUW20,UWY}]
	\label{PlanarProduct}
	Every planar graph is a subgraph of:
	\begin{compactenum}[(a)]
		\item $H\boxtimes P$ for some graph $H$ of treewidth at most $8$ and for some path $P$,
		\item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

% \note{DW}{In 	\cref{PlanarProduct} and elsewhere, we really should write, ``Every planar graph is isomorphic to a subgraph of''. One alternative is to define that $H$ is \defin{contained} in $G$ if $H$ is isomorphic to a subgraph of $G$. I suggest the latter.}

% \note{DW}{I suggest we replace $8$ by $6$ here, and replace ``The starting point for this work is the following `product structure theorem' for planar graphs by \citet{DJMMUW20} (improved by \citet{UWY})''. Okay?}

Here $\boxtimes$ is the strong product,\!\footnote{The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if
	$v=w$ and $xy\in E(B)$, or
	$x=y$ and $vw\in E(A)$, or
	$vw\in E(A)$ and $xy\in E(B)$.}
and treewidth\footnote{For a tree $T$, a \defin{$T$-decomposition} of a graph $G$ is a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $T$ such that
(i) for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and
(ii) for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected. The \defin{width} of $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  A \defin{tree-decomposition} is a $T$-decomposition for any tree $T$. The \defin{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.} is an invariant that measures how `tree-like' a given graph is; see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarProduct} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded treewidth to be generalised for planar graphs (with different constants).

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).
\item \citet{DFMS21} use it to make dramatic improvements to the best known bounds for $p$-centered colourings of planar graphs.
\item \citet{bonamy.gavoille.ea:shorter} use it to find shorter adjacency labellings of planar graphs (improving on a sequence of results going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}), and \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs.
\item The result of \citet{DEJGMM21} implies that, for every integer
$n>0$, there is a `induced-universal' graph $U_n$ with $n^{1+o(1)}$ vertices such that every $n$-vertex planar graph is an induced
subgraph of $U_n$. This has recently been extended by \citet{EJM}, who show the existence of a `subgraph-universal' graph with $(1+\epsilon)n$ vertices that contains every $n$-vertex planar graph as a subgraph as well as the existence of an `induced universal' graph with $n^{1+o(1)}$ vertices \emph{and} edges.  The former result is the first progress on subgraph-universal graphs for planar graphs since the $O(n^{3/2})$-vertex construction of \citet{babai.chung.ea:on}.
% aking the first progress on this problem since the work of \citet{babai.xx} in 1984.
\end{compactitem}

\note{DW}{Delete `from 1992' and  `from 2007'? replace ``going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}'' by
	``starting with the work of \citet{kannan.naor.ea:implicit}''}
\note{PM}{Was this a referee request, or just because we don't need to sell it anymore?  I'm fine with the change either way, but didn't implement it.}

% \note{DW}{Add \citep{EJM} to the story.}

@@

\note{PM}{Kind of, but I suggest we keep that paragraph and section anyway. Justification: They're both short and they describe the exact bounds for bounded genus graphs, which you don't get from the one-paragraph description in Section~1.0.  A}

@@

\note{DW}{Use $\SS$ for shortcut system, okay?}
\note{PM}{Ok.}

@@

\begin{thm}[\citep{DJMMUW20,UWY,DHHW}]
	\label{GenusProduct}
	Every graph of Euler genus $g$ is a subgraph of:
	\begin{compactenum}[(a)]
		\item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
		\item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

\note{DW}{I suggest we replace $2g+8$ by $2g+6$ here and replace 4 by 3, and cite \citep{DHHW,UWY}. This will improve the bounds and (more importantly) simplify the presentation of various results (such as \cref{StringPartition,PowerGenus}) which currently distinguish the $g=0$ case. }
\note{PM}{I agree with this suggestion.  I've only added the other two references here, and haven't changed anything elsewhere yet.}

@@

  Next, remove the node $x_0$ from $T_0$, and consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.
  % To see that $(B_x:x\in V(T))$ satisifies \tref{ancestor-edge}, observe that,
  % if $xy\in E(H)$, then at least one of $x$ or $y$ is contained in $B_z$ for every node $z$ on the path from $x$ to $y$ in $T$.  If neither $x$ nor $y$ is an ancestor of the other, then some node $z$ on this path has $T$-depth less than that of $x$ and $y$.  If $x\in B_z$ this contradicts the fact that $x$ is the root of $T[x]$.  If $y\in B_z$ this contradicts the fact that $y$ is the root of $T[y]$.
\end{proof}

% \referee{1}{6. page 6, last paragraph of Lemma 2: This isn’t quite right since you have to be careful with $x_0$.}
% \note{PM}{I addressed this by removing $x_0$ from $T_0$ at the beginning of the second paragraph.}

@@

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.\footnote{The sets in $\mathcal{S}$ are disjoint because $a:V(G)\to V(T)$ is a function.  The sets in $\mathcal{S}$ cover $V(G)$ since, for each $v\in V(G)$, $v\in S_{a(v)}$.} Let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

\referee{2}{Page 8.  Can you add a brief explanation why $\mathcal{S}$ is a
	partition of $V(G)$?  It is clear that the sets in $\mathcal{S}$ are
	disjoint, but why do they cover $V(G)$?}
\note{PM}{I added an explanation in a footnote since this is obvious.}
